---
title: "Scripts to find correlation coefficients"
author: "your name"
date: "the date"
output:
  html_document:
    df_print: paged
---

```{r, include = FALSE}
# you can ignore this chunk - it just makes the rendered script look nicer
knitr::opts_chunk$set(message=FALSE, warning=FALSE)
```


This notebook contains three examples of use of correlation tests on continuous data.

You can copy whichever parts meets your needs, delete the rest and adapt it to your data.


## Example of normally distributed data - no evidence of correlation


**Question**: Is tree height correlated with tree diameter at breast height?

### load packages
```{r}
library(tidyverse)
library(here)
library(cowplot)
```

### Load data
Here we will choose tree data. Your data file should be a csv file, inside the data subolder of your project

```{r}
filepath <- here("data","tree_vol.csv")
trees <- read_csv(filepath)
glimpse(trees)
```

### Plot the data

```{r}
trees |>
  ggplot(aes(x = diameter, y = height)) + # what we want to plot
  geom_point() + # what kind of plot
  labs(x = "Tree diameter (m)",
       y = "Tree height (m)") + # axis labels
  theme_cowplot() # give the plot a nice 'look'
```

Does it look as though tree diameter and height are linearly correlated?


#### Check for normality
```{r}
shapiro.test(trees$diameter) 
shapiro.test(trees$height)
```

If p > 0.05 in both cases, then we can assume normality 

If both are distributed normally, then we check for Pearson's r, if not we check for Spearman's Rank r

### Find correlation coefficient
```{r}
cor.test(trees$diameter,trees$height, method = "pearson")
```

#### Interpret the output

* What is r?  
* Is p < 0.05? 
* Does the confidence interval encompass zero?
* Is r significantly different from zero? 


## Example of normally distributed data - where there is evidence of correlation


**Question**: Is virginica petal width correlated with virginica petal length?

### Load packages
```{r}
library(tidyverse)
library(here)
library(cowplot)
```

### Load data
Here we choose the iris data set
```{r}
filepath <- here("data","iris.csv")
iris <- read_csv(filepath)
glimpse(iris)
```

### Select subset of data
Let us select the data for the setosa species
```{r}
virginica <- iris %>%
  filter(Species == "virginica")
```

### Plot the data - petal length vs petal width

```{r}
virginica %>%
  ggplot(aes(x = Petal.Width, y = Petal.Length)) + # what we want to plot
  geom_point() + # what kind of plot
  labs(x = "Petal diameter (cm)",
       y = "Petal width (cm)") + # axis labels
  theme_cowplot() # give the plot a nice 'look'
```


Does it look as though petal width and length are linearly correlated?


### Check for normality
```{r}
shapiro.test(virginica$Petal.Length)  # p > 0.05?
shapiro.test(virginica$Petal.Width) # p > 0.05?
```


### Find correlation coefficient - pearson's r
```{r}
cor.test(virginica$Petal.Length,virginica$Petal.Width, method = "pearson")
```

#### Interpret the output

* What is r?  
* Is p < 0.05? 
* Does the confidence interval encompass zero?
* Is r significantly different from zero?

## Example of correlation test where we have to use Spearman's rank r~sp~


**Question**: Is WEMWBS score correlated with nep score?

### Load packages
```{r}
library(tidyverse)
library(here)
library(cowplot)
```

### Load data
We choose the marine wellness data set - where the data are ordinal data
```{r}
filepath <- here("data","marine_wellness.csv")
wellness <- read_csv(filepath)
glimpse(wellness)
```

We have wellness scores for 288 individuals according to two scales - WEMWBS and nep

### Plot the data - WEMWBS score vs nep score
```{r}
wellness %>%
  ggplot(aes(x = WEMWBS, y = nep)) + # what we want to plot
  geom_point() + # what kind of plot
  labs(x = "WEMWBS score",
       y = "nep score") + # axis labels
  theme_cowplot() # give the plot a nice 'look'
```

Does it look as though the relationship between the two scores is monotonic?

### Test for correlation
```{r}
cor.test(wellness$WEMWBS,wellness$nep, method = "spearman",,exact = FALSE)
```

#### Interpret the output

* What is r~sp~?  
* Is p < 0.05? 
* Is r significantly different from zero?
