---
title: "Example Analyses"
author: "Michael Hunt"
date: "16/05/2024"
output:
  html_document:
    df_print: paged
---

This notebook uses the penguin data set from the palmerpenguins package to illustrate the use of correlation, linear regression and t-test / Wilcox tests in R.

```{r}
library(tidyverse)
library(palmerpenguins)
library(cowplot)
library(rstatix)
```

```{r}
glimpse(penguins)
```

```{r}
penguin_adelie <- penguins |> 
  filter(species == "penguin_adelie")
glimpse(penguin_adelie)
```

### Correlation

```{r}
penguin_adelie |> 
  ggplot(aes(x=bill_length_mm, bill_depth_mm)) +
  geom_point() +
  labs( x = "Bill length (mm)",
        y = "Bill depth (mm)") +
  theme_cowplot()
```

```{r}
bill_length<-penguin_adelie$bill_length_mm
bill_depth<-penguin_adelie$bill_depth_mm

cor.test(bill_length, bill_depth, method = "pearson")
```


### Simple linear regression
```{r}
bill.model<-lm(bill_depth ~ bill_length, data = penguin_adelie)
anova(bill.model)
summary(bill.model)
```

```{r}
penguin_adelie |> 
  ggplot(aes(x=bill_length_mm, bill_depth_mm)) +
  geom_point() +
  geom_smooth(method="lm", linewidth = 0.5) +
  labs( x = "Bill length (mm)",
        y = "Bill depth (mm)") +
  theme_cowplot() +
  theme(panel.grid = element_line(color = "#8ccde3",
                                  linetype = 2,
                                  linewidth=0.5))
```

```{r}
penguin_adelie |> 
  ggplot(aes(x=bill_length_mm, bill_depth_mm)) +
  geom_point() +
  geom_smooth(method="lm", linewidth = 0.5) +
  labs( x = "Bill length (mm)",
        y = "Bill depth (mm)") +
  xlim(0,46) +
  ylim(0,24) +
  theme_cowplot() +
  theme(panel.grid = element_line(color = "#8ccde3",
                                  linetype = 2,
                                  linewidth=0.5))
```


### t-test / Wilcox tests for difference

```{r}
penguins |>
  ggplot(aes(x=species,y=body_mass_g)) +
  geom_boxplot() +
  labs(x= " Species",
       y = "Body mass (g)") +
  theme_cowplot()
```

```{r}
penguins |>
  ggplot(aes(x=body_mass_g)) +
  geom_histogram() +
  labs(x= " Body mass (g)",
       y = "Count") +
  facet_wrap(~ species, ncol = 1) +
  theme_cowplot()
```
  
  

#### test for normality
```{r}
penguins |>
  group_by(species) |>
  shapiro_test(body_mass_g)
```

#### t_test / Wilcox test
```{r}
penguin_adelie_bodymass <- penguins |> filter(species == "penguin_adelie") |> select(body_mass_g) |> pull()
chinstrap_bodymass <- penguins |> filter(species == "Chinstrap") |> select(body_mass_g) |> pull()
gentoo_bodymass <- penguins |> filter(species == "Gentoo") |> select(body_mass_g) |> pull()
```

Do penguin_adelie penguin body masses differ from those of Chinstrap penguins?
```{r}
t.test(penguin_adelie_bodymass,chinstrap_bodymass)
wilcox.test(penguin_adelie_bodymass,chinstrap_bodymass)
```

Do Chinstrap penguin body masses differ from those of Gentoo penguins?
```{r}
t.test(chinstrap_bodymass,gentoo_bodymass)
```
