---
title: "Generalised linear models"
author: "Michael Hunt"
date: "23-01-2024"
output:
  html_document:
    df_print: paged
---

```{r set up, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```


This exercise is taken from Chapter 7 of Beckerman, Childs and Petchey: Getting Started with R

We study variation of number of offspring of Soay sheep with mother's body weight. These are count data. They are not normally distributed since they are discrete and bounded by zero - a ewe cannot have a negative number of births. Moreover, the variance of the data set is not constant - the spread of values tends to increase the more offspring a ewe has. This is common (but not universal) with count data.



### Load packages
```{r load packages}
library(tidyverse)
library(ggfortify) 
library(here)
library(cowplot)
```

### Load sheep data
```{r load data}
filepath <- here("data","soay_sheep_fitness.csv")
soay<-read_csv(filepath)

# check its structure
glimpse(soay)
```


### Make the first plot to explore the data
```{r first plot}
ggplot(soay,aes(x=body.size,y=fitness))+
  geom_point()+
  geom_smooth(method='lm',se=FALSE, linewidth = 0.5)+
  xlab('Body mass (kg)')+
  ylab('Lifetime fitness')+
  theme_cowplot()
```

A few things stand out from this plot:

* There is a clear trend: as the ewe's body mass increases, so does the number of offspring tend to increase. That makes sense: a heavier ewe would have more resources to support one lamb after another.
* The trend is *not* linear. There is a clear upwards curve.
* The spread of values increases as the mother's body mass increases. This is common with count data.
* There are no negative response values - while a eewe could have no lambs, it definitely cannot have a negative numer of lambs.

### The wrong way to analyse the data - do a linear model

We should have an inkling that this is the wrong way since the variation of offspring with body weight is clearly not linear!
FOr insight, let's do it anyway using lm(), and look at the diagnostic plots we get when we give the resulting model `soay.linear` to `autoplot()`

```{r wrong analysis}
soay.linear<-lm(fitness~body.size,data=soay)
autoplot(soay.linear,smooth.colour=NA) + theme_cowplot()
```


### The right way - using a GLM
```{r right analysis with GLM}
soay.glm<-glm(fitness~body.size,data=soay,family=poisson(link=log))
autoplot(soay.glm,smooth.colour=NA) + theme_cowplot()
```


### Interpret the GLM
```{r interpret model}
# use anova() and summary() to interpret the model
anova(soay.glm) # sequential sums of squares table
summary(soay.glm) # coefficients table
```

### Remake the figure and add fitted line

#### Find min and max of the x data
```{r min and max}
min.size<-min(soay$body.size)
max.size<-max(soay$body.size)
```

#### Make the new.x values.
We use the 'body.size' variable name to name the column just as it is in the orginal data
```{r}
new.x<-expand.grid(body.size=seq(min.size,max.size,length.out=1000))
```

#### Generate fits and standard errors
```{r}
new.y<-predict(soay.glm,newdata=new.x,se.fit=TRUE)
new.y<-data.frame(new.y)
head(new.y)
```

#### Housekeeping to bring new.x and new.y together
```{r}
addThese<-data.frame(new.x,new.y)
addThese<-rename(addThese,fitness=fit)
head(addThese)
```

#### Calculate confidence intervals and include them in addThese
```{r}
addThese<-mutate(addThese,
                 lwr=fitness-1.96*se.fit,
                 upr=fitness+1.96*se.fit)
```

#### Plot the final figure
```{r}
ggplot(soay,aes(x=body.size,y=fitness)) +
  # first show the raw data
  geom_point(size=3,alpha=0.5)+
  # now add the fit and confidence intervals
  # we don't need to specify body.size and fitness as they are inherited from above
  geom_smooth(data=addThese,aes(ymin=lwr,ymax=upr),stat='identity')+
  theme_cowplot()
```

This did not work! We need to account for the link function. For a poisson model, this is the log(). Hence, the predictions are the log of the expected fitness. Here is the fix:

#### Start again with addThese:
```{r}
addThese<-data.frame(new.x,new.y)
# then:
addThese<-mutate(addThese,
                 fitness=exp(fit),
                 lwr=exp(fit-1.96*se.fit),
                 upr=exp(fit+1.96*se.fit))
head(addThese)
```

#### Redo the final plot
```{r}
ggplot(soay,aes(x=body.size,y=fitness)) +
  # first show the raw data
  geom_point(size=3,alpha=0.5)+
  # now add the fit and confidence intervals
  # we don't need to specify body.size and fitness as they are inherited from above
  geom_smooth()+
  geom_smooth(data = addThese,aes(ymin = lwr,ymax = upr), stat = 'identity')+
  theme_cowplot()
```
