---
title: "Maths notation in R titles, labels and annotations for ggplot2"
author: "Michael Hunt"
date: "18/12/2024"
output:
  html_document:
    df_print: paged
---

```{r, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r, echo = FALSE}
library(tidyverse)
```



Often in plots one needs to use mathematical expressions, suffixes/superscripts, greeek letters or other unusual formatting.

Here we show one way of doing this, using `bquote()`. Over the years I have found this to be the simplest way.

### Use `bquote()`

The four rules

* Strings – Require quotes wrapped w/ tilde separator (e.g., "my text"  ~).
* Math Expressions – Unquoted & follow `?plotmath`
* Numbers – Unquoted when part of math notation
* Variables – Use .() (pass in string or numeric)

Examples of all of these are shown in the labels, title and annotations included in the plot below:


```{r}
df <- tibble(x=1:10,y=1:10)
cor <- 0.456
df |>
  ggplot(aes(x=x,y=y)) +
  geom_point() +
  labs( x = bquote("An axis label with suffix and superscript:" ~ x[i]^2), 
        y = bquote("An axis label with greek letters:" ~ alpha ~ Beta ~ mu ~ m),
        title = bquote("Hello"~ r[xy] == .(cor) ~ "and" ~ CO[2] ~ "and" ~ B^2 ~ "and" ~ m^{-1})) +
  annotate(geom="text", x = 5, y = 7, label =  bquote("Hello" ~ r[xy] == 0.678 ~ "and" ~ B^2)) +
  annotate(geom="text", x = 7, y = 9, label =  bquote("A big annotation"), size = 12) +
  annotate(geom="text", x = 7, y = 2, label =  bquote("A red annotation"), colour = "red") +
  theme_classic()
```

### ?plotmath

`plotmath` expressions can be used for mathematical annotation in text within plots in R when writing titles, subtitles, axis labels, legends and annotations. It works in both base R graphics and ggplot2.

In the console pane, type `?plotmath` in the console pane to see the full list of options.

Some key rules are:

* subscripts: O[2] gives O~2~
* superscripts: m^2 gives m^2^, m^{-1} gives m^-1^
* Lower case Greek: alpha, beta etc gives $\alpha, \beta$ etc, so mu
* Upper case Greek: Delta, Gamma etc gives $\Delta, \Gamma$ etc


