---
title: "Analysis of ozone data from two parts of a city"
author: "Michael Hunt"
date: "13-11-2024"
output:
  html_document:
    df_print: paged
---

```{r, include = FALSE}
# This chunk makes the rendered version look prettier
# it supresses warnings and other messages
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This uses the parametric t-test for a difference between two samples

### load the packages we need
```{r}
library(tidyverse)
library(here)
```


### load the data

This data set is saved as a .csv file, so we use the `read_csv()` function to open it.
```{r}
filepath <- here("data","ozone.csv")
ozone<-read_csv(filepath)
glimpse(ozone)
```
### summarise the data
```{r}
ozone.summary<-ozone |>
group_by(garden.location) |>
summarise(n = n(),
          mean.ozone = mean(ozone),
          se.ozone = sd(ozone)/sqrt(n()))
ozone.summary
```

### plot histograms for each location
```{r}
ozone |>
ggplot(aes(x=ozone))+
  geom_histogram(binwidth=10,fill="darkred")+
  facet_wrap(~garden.location,ncol=1) +
  theme_classic()
```

### plot box plot for the two locations
```{r}
ozone |>
  ggplot(aes(x=garden.location,y=ozone))+
  geom_boxplot()+
  labs(x="Garden Location",
       y="Ozone concentration (ppb)") +
  theme_classic()

```


# are the ozone levels in each location normally distributed?
```{r}
ozone |>
  group_by(garden.location) |>
  summarise(sw=shapiro.test(ozone)$p.value)
```

### graphical check for normality:

#### quantile-quantile plots for ozone levels in each location
```{r}
ozone |>
  ggplot(aes(sample=ozone)) +
  stat_qq(colour="blue") +
  stat_qq_line() +
  facet_wrap(~garden.location) +
  theme_classic()
```

### and finally, the actual _t_-test
```{r}
t.test(ozone~garden.location,data=ozone)
```
